Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy

### 2025-06-04 Text/Articles/RENYI_ENTROPY_SAR_2025/Submissions Files/R1/Responses/Responses-Associate-Editor.tex

\vspace{0.5em}

\textbf{Equation (5): Rényi entropy of the $\mathcal{G}^{0}_{\!I}$ model.}  
This expression presents the closed-form Rényi entropy of the heterogeneous $\mathcal{G}^{0}_{\!I}$ distribution as the sum of two components:
\[
H_\lambda\bigl(\mathcal{G}^{0}_{\!I}\bigr)
=\;H_\lambda\bigl(\Gamma_{\text{SAR}}\bigr)
\;+\;\bigl[\text{extra\ term depending on }\alpha\bigr].
\]

In intuitive terms:
\begin{itemize}
	\item $H_\lambda(\Gamma_{\text{SAR}})$ represents the “baseline entropy” of fully developed speckle (homogeneous clutter).
	\item The additional term captures how much entropy increases due to texture (heterogeneity), caused by the roughness parameter $\alpha$.
\end{itemize}
When $\alpha\!\to\!-\infty$ (fully developed speckle), the excess term vanishes and (5) collapses to the homogeneous case in (4).  This motivates our test: we check whether the \emph{excess} is statistically different from zero.

\vspace{0.3em}
\begin{quote}

\textcolor{blue}{To improve clarity, we expanded the explanation following Equation~(5) and clarified that Figure~1 illustrates the behavior of the excess entropy term as $\alpha$ decreases, confirming its convergence to the homogeneous case.}

\end{quote}
\end{responsebox}

@@

\begin{responsebox}{}
\vspace{0.5em}
\textbf{Equation (8): test statistic.}  
The test statistic is simply defined as:
\[
S_{\widetilde H_\lambda}(Z;L)
= \text{“estimated entropy”} - \text{“theoretical entropy under } H_0\text{”}.
\]
Its purpose is to detect deviations from homogeneity:
\begin{itemize}
	\item  If the data are homogeneous, both terms match closely $\;\Rightarrow\;S\approx0$.
	\item  If the data are heterogeneous, the estimated entropy exceeds the theoretical one $\;\Rightarrow\;|S|$ increases and the $p$-value becomes small.
\end{itemize}

This formulation avoids parameter estimation for the $\mathcal{G}^0_I$ model, offering a simple, interpretable, and statistically grounded test.

\begin{quote}
\textcolor{blue}{We added a brief explanation below Equation (8), and clarified in the main text (related to Figure~4) that the centered distribution confirms correct behavior under $H_0$.}

\textbf{Why Eq.(5) benefits practitioners.}  
Closed-form Rényi entropy for $\mathcal{G}^{0}_{\!I}$ is absent from the SAR literature; Having it allows others to incorporate an analytical feature into algorithms and applications.
\begin{quote}
\textcolor{blue}{These additions are included in Sections II-B and III-D of the revised manuscript.}

\textcolor{blue}{Note: Due to space limitations, the derivation of Equation~(5), which was previously included in the Appendix, has been omitted in the revised version.}
\end{quote}
\end{responsebox}

\vspace{1em}
\begin{reviewbox}{Comment 2}
2. The current methodology assumes that the number of looks, L, is known a priori. In practice, this parameter may be unknown or inaccurately estimated from SAR metadata. The authors should discuss the sensitivity of the proposed test to variations or inaccuracies in L, and optionally suggest techniques for estimating L directly from the data or include robustness analysis. 
\end{reviewbox}

@@

\end{quote}

\end{responsebox}

\begin{reviewbox}{Comment 3}
3. Figures (Figs. 6–8) are informative but could benefit from quantitative evaluation. Add comparative metrics beyond visual comparison.
\end{reviewbox}

\begin{responsebox}{Response}
We thank the reviewer for the suggestion. We have added a quantitative performance evaluation to the revised manuscript to complement the visual comparison.

Specifically, we manually selected eight regions of interest (ROIs) in each SAR image: four clearly homogeneous areas (e.g., water, croplands) and four clearly heterogeneous areas (e.g., urban centers). Each ROI was labeled as class 0 (blue) or class 1 (red), as illustrated in Figs.~5a, 6a, and 7a (corresponding to Figs.~6a, 7a, and 8a in the original submission).

These ROIs were rasterized to create a partial ground truth mask. Binary decision maps were obtained by thresholding the $p$-value maps at a 0.05 significance level, assigning a value of 1 to pixels with $p < 0.05$ (heterogeneous), and 0 otherwise (homogeneous). These predictions were compared to the ground truth labels, and standard metrics—F1-score, kappa coefficient, and overall accuracy—were computed. The results are reported in Table~III.

To further support the evaluation, we computed ROC curves and the corresponding AUC values for each method, as shown in Fig.~9.

\begin{quote}
\textcolor{blue}{These additions are included in Section~IV-C of the revised manuscript.}
\end{quote}
\end{responsebox}

### 2025-06-04 Text/Articles/RENYI_ENTROPY_SAR_2025/Submissions Files/R1/Responses/Responses-Reviewer2.tex

\vspace{0.5cm}
\noindent Dear Editors
\bigskip

\noindent We are grateful to the reviewers for the valuable comments and the time spent on checking our manuscript. 
We have addressed their comments towards improving the paper contents and presentation. 
Below we detail the changes and updates incorporated into the manuscript in this round of review.

\medskip

%-------------------------------------
\vspace{2em}
\begin{reviewbox}{Comment 1}
1. The language needs further improvement and should be reviewed by a language expert.
\end{reviewbox}

\begin{responsebox}{Response}
We have carefully revised the manuscript to improve its clarity.

@@

\end{responsebox}

\vspace{3em}
\begin{reviewbox}{Comment 2}
2. The scientific content of the article is weak and requires further expansion.
\end{reviewbox}

\begin{responsebox}{Response}

In the revised manuscript, we have expanded several sections to reinforce the scientific content and clarify the contributions of our work. 
This article presents a complete statistical framework for detecting heterogeneity in SAR images using a Rényi entropy-based test. 
The proposed method is analytically grounded, unsupervised, and provides interpretable $p$-values. 
We enhance a non-parametric entropy estimator via bootstrap correction, validated its statistical properties through a Monte Carlo study, and demonstrated its application on real SAR data under challenging conditions. 
The work further includes quantitative evaluation using standard metrics and ROC curves. We believe these contributions represent both theoretical and practical advances in SAR image analysis.

@@

\end{responsebox}

\vspace{10em}
\begin{reviewbox}{Comment 3}
 3. I did not observe any comparison of the authors' work's novelty index with the latest articles published in the field for the current years.
\end{reviewbox}

\begin{responsebox}{Response}
We appreciate the reviewer’s suggestion. In response, we have expanded the introduction to better contextualize our contribution relative to recent work.

Specifically, we now cite recent studies that explore entropy-based methods for SAR analysis and classification. For example, Cassetti et al. (2022) evaluated several Shannon entropy estimators in both supervised and unsupervised SAR classification. Gallet et al. (2024) proposed an explainable classification framework based on learned Rényi divergences, demonstrating robustness under label noise and data scarcity. Parikh (2019) reviewed deep learning approaches for SAR classification and identified key challenges, including the need for large labeled datasets and complex hyperparameter tuning.

In contrast to these data-driven or classification-focused methods, our work addresses heterogeneity detection from a statistically grounded perspective. We propose a hypothesis test based on a non-parametric Rényi entropy estimator, enhanced via bootstrap bias correction. This approach does not require training data, produces interpretable $p$-values, and is particularly suited for unsupervised analysis and scenarios with limited ground-truth information.

\begin{quote}
\textcolor{blue}{These comparisons and citations have been added to the revised Introduction to clarify the novelty and positioning of our contribution.}
\end{quote}
\end{responsebox}

\vspace{3em}

\begin{reviewbox}{Comment 4}
4. Conclusion requires redrafting.
\end{reviewbox}
\begin{responsebox}{Response}
The conclusion section has been fully revised to improve clarity and better reflect the main contributions, evaluation results, and future research directions of the study.

### 2025-06-04 Text/Articles/RENYI_ENTROPY_SAR_2025/R1-Renyi-entropy-sar.qmd
Close to finishing

      Data and code are available at: <https://github.com/rjaneth/renyi-entropy> 
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Quantifying surface roughness in synthetic aperture radar (SAR) data is critical for accurate geophysical interpretation and remote sensing applications. 
  We propose a test statistic based on a non-parametric estimation of Rényi entropy to characterize surface roughness from SAR intensity data. 
  The statistic is refined using bootstrap to improve its stability, size, and power. 
  This approach enhances roughness quantification by capturing scale-dependent variations and addressing data-driven uncertainty.
  Experimental results demonstrate the robustness of the proposed method in distinguishing roughness patterns.
  <!-- , offering a statistically rigorous tool for SAR-based terrain analysis. -->
  <!-- Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems.  -->
  <!-- A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes.  -->
  <!-- In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed.  -->
  <!-- In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task.  -->
  <!-- In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy.  -->
  <!-- In particular, we use a non-parametric estimator for the Rényi entropy and investigate some of its properties.  -->
  <!-- As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories.  -->
  <!-- The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy. -->
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR, hypothesis tests, bootstrap]

@@

## The Proposed Test
The test statistic quantifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0}\hspace{-0.5em},
$$
the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids the need to estimate $\mathcal{G}^0_I$ parameters  such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.

>>>>

      This research, including its outputs, e.g., this manuscript, was executed in Quarto and is fully reproducible. We used RStudio version 2024.12.1+563, and R version 4.4.2. Data and code are available at: <https://github.com/rjaneth/renyi-entropy> 
    # bio: |
    #   Use `IEEEbiographynophoto` and the author name
    #   as the argument followed by the biography text.
    # note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
abstract: |
  Quantifying surface roughness in synthetic aperture radar (SAR) data is critical for accurate geophysical interpretation and remote sensing applications. 
  We propose a test statistic based on a non-parametric estimation of Rényi entropy to characterize surface roughness from SAR intensity data. 
  The statistic is refined using bootstrap to improve its stability, size, and power. 
  This approach enhances roughness quantification by capturing scale-dependent variations and addressing data-driven uncertainty.
  Experimental results demonstrate the robustness of the proposed method in distinguishing roughness patterns.
  <!-- , offering a statistically rigorous tool for SAR-based terrain analysis. -->
  <!-- Synthetic aperture radar (SAR) systems have already been successfully used to solve remote sensing problems.  -->
  <!-- A disadvantage of SAR images is the presence of speckle, which is fully developed in homogeneous areas and gamma-distributed in these scenes.  -->
  <!-- In heterogeneous areas, the intensity values are $\mathcal{G}^0_I$-distributed.  -->
  <!-- In this way, the identification of roughness SAR areas (as opposed to homogeneous ones) is an important task.  -->
  <!-- In this work, we propose a family of hypothesis tests driven by an order parameter to identify roughness features in SAR intensity data using Rényi entropy.  -->
  <!-- In particular, we use a non-parametric estimator for the Rényi entropy and investigate some of its properties.  -->
  <!-- As a practical evaluation method, we develop $p$-value maps on which one can observe both (i) the heterogeneous evidence change per texture and (ii) the prediction of homogeneous and heterogeneous categories.  -->
  <!-- The results are in favor of the Rényi-based heterogeneity detector compared to the one based on Shannon entropy. -->
keywords: [Rényi entropy, Gamma distribution, heterogeneity, SAR, hypothesis tests, bootstrap]

@@

## The Proposed Test

We propose a test statistic that identifies the discrepancy between estimated and theoretical entropy under homogeneity. Since $L\geq1$ is known, we define the test statistic as follows:
\begin{multline}
\label{eq-test}
S_{\widetilde{H}_{\lambda}}(\bm{Z}; L) = \widetilde{H}_{\lambda} - \bigl\{\ln \widehat{\mu} - \ln L + \frac{1}{1-\lambda}
\bigl[-\lambda\,\ln\Gamma(L) \\ 
+ \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  
- \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\bigr]\bigr\},
\end{multline}
where $\widehat{\mu}={n}^{-1}\sum_{i=1}^n Z_{i}$ is the sample mean.
The test statistic can be interpreted as:
$$
S_{\widetilde H_\lambda} 
= 
\underbrace{\widetilde H_\lambda}_{\text{estimated}} 
\;-\;
\underbrace{H_\lambda\bigl(\Gamma_{\mathrm{SAR}}\bigr)}_{\text{theoretical under } H_0}\hspace{-0.5em},
$$
the difference between the estimated entropy and the expected value under homogeneity. Values close to zero indicate that the region behaves like fully developed speckle, while large positive values signal excess entropy and, thus, heterogeneity.
This formulation avoids the need to estimate $\mathcal{G}^0_I$ parameters  such as $\alpha$, offering a simple, interpretable, and statistically grounded test. Moreover, the method remains effective for small samples, aided by the bootstrap bias correction.
